Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-03-20 paper/main.tex
typo
changes in experiments (text)

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$\rho=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\ldots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given a
set of states $S\subseteq Q$, let $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denote the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m=\{t\in\delta\mid m\in
\mks(t)\}$ denote the set of all transitions
marked with $m$.
% \todo{The sets $\delta_m$ are called \emph{acceptance sets}. Given
% an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
% $\delta_{S,m}=\delta_S\cap\delta_m$.}
A set of transitions $T\subseteq\delta$ is called a \emph{cycle} if
there exists a path from a state $p$ to the same state containing each
transition of $T$ at least once and no transition outside
$T$. Finally, we assume that each TELA $\mA$ contains only states $q$
that are reachable from the initial state $q_I$ (i.e., $q=q_I$ or
there is a path from $q_I$ to $q$) as states that are not reachable
from $q_I$ can be eliminated without any impact on $L(\mA)$. We also
assume that each TELA has at least one SCC as automata without any SCC
trivially desribe an empty language.

@@

To analyze the length of the formulas, we start with its subformulas.
One can easily check that
$|\msat_\varphi(\vec{e})|\in\bigO(|\varphi|\cdot|\delta|)$ and
$|\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})|\in\bigO(C\cdot
K\cdot|\delta|)$. Further,
$|\mcycle_1(\vec{e})|,|\mcycle_2(\vec{e})|\in\bigO(S\cdot|\delta|)$, % and
%$|\mcycle_2(\vec{e})|\in\bigO(S\cdot|\delta|+|\delta|)=\bigO(S\cdot|\delta|)$,
where $S$ is the number of SCCs in the automaton.  Next,
$|\mcycle_3(\vec{e})|\in\bigO(|\delta|+|Q|)$, which can be simplified
to $|\mcycle_3(\vec{e})|\in\bigO(|\delta|)$ as $|Q|\le|\delta|$
follows from the assumptions that all states are reachable and each
automaton has at least one SCC. Altogether, we get
$|\Phi_{1,C,K}|,|\Phi_{2,C,K}|\in\bigO(S\cdot|\delta|+|\varphi|\cdot|\delta|+C\cdot
K\cdot|\delta|)=\bigO\big((S+|\varphi|+C\cdot K)\cdot|\delta|\big)$
and
$|\Phi_{3,C,K}|\in\bigO(|\delta|+|\varphi|\cdot|\delta|+C\cdot
K\cdot|\delta|)=\bigO\big((|\varphi|+C\cdot K)\cdot|\delta|\big)$.

@@

Finally, we mention two simple optimizations of the formula
construction, which are always applied in the rest of the paper.

@@

@@

\todo{Zminit, ze nase QBF formule navic rikaji, ze v zadne
  klauzulu/cude neni zaroven $\Fin m$ a $\Inf m$ a ze cela formule
  ma alepson 1 $\Inf$ nebo $\Fin$? Je to tak, Terko?}

@@

\todo{Nekde zminit, ze inkrementalni pristup je implementovan jen pro prvni while loop? Je to tak, Terko?}

If some call of the function
$\mathit{satisfiable}(\Phi_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In the case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance marks considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
did not decide within the time limit. If $X$ is \texttt{T}, a longer
timeout may lead to further reductions. If the multi-level reduction
is used, the automaton name contains the information from all
levels. For example, the name `\texttt{L1\_5\_U L2\_3\_U L3\_1\_T}'
means that level 1 reduced the number of marks to 6 (reduction to 5 is
impossible on this level), level 2 reduced it to 4, and level 3 to 2
as the QBF solver did not finish in the time limit when trying to
reduce the number of marks to
1.%when tryint the automaton may be potentially reduced further.

@@

\todo{Upravit experimenty: pridat dalsi 2 sady automatu.}
We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. We used the tool
\texttt{ltlfilt} of the Spot library to simplify the formulas and
remove duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf}, and \texttt{ltl2dgra} (known as
Rabinzer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four automata sets
produced by the individual tools contains less than 348 automata as not
all of the formulas are successfully translated due to a timeout of 60
seconds. Further, we have removed automata with 0 or 1 acceptance
marks as there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators, and the number of automata produced by each
translator after removing the automata with less than two marks.

\begin{table}[t]
\caption{LTL to automata translators used for the experimental evaluation
and the number of successfully translated formulas.}
\label{tab:ltltranslators}
\centering
% \begin{tabular}{lr@{~~~~~~}lr}
% \toprule
% tool        & version&homepage & ~~~automata \\
% \midrule
% \texttt{ltl2dela}~~~ & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 129\\
% \texttt{ltl2dgra} & 21.0 & & 234\\
% \texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 70\\
% \texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 91\\
% \bottomrule
% \end{tabular}
%
\begin{tabular}{lrrrr}
  \toprule
  &&& \multicolumn{2}{c}{constructed automata with} \\ \cmidrule(lr){4-5}
  tool &\!\!\!version/homepage & timeout & 0 or 1 mark & \textbf{at least 2 marks}\\
  \midrule
  \texttt{ltl2dela} & 21.0\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
       & ? & ? & \textbf{129} \\
  \texttt{ltl2dgra} & 21.0\footnotemark[1]
       & ? & ? & \textbf{234}\\
  \texttt{ltl2tgba} & 2.10.4\tablefootnote{\url{https://spot.lrde.epita.fr}}
       & ? & ? & \textbf{70}\\
  \texttt{ltl3tela} & 2.2.0\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
       & ? & ? & \textbf{91}\\
  \bottomrule
\end{tabular}
\end{table}

>>>>

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$\rho=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\ldots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given a
set of states $S\subseteq Q$, let $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denote the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m=\{t\in\delta\mid m\in
\mks(t)\}$ denote the set of all transitions
marked with $m$.
% \todo{The sets $\delta_m$ are called \emph{acceptance sets}. Given
% an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
% $\delta_{S,m}=\delta_S\cap\delta_m$.}
A set of transitions $T\subseteq\delta$ is called a \emph{cycle} if
there exists a path from a state $p$ to the same state containing each
transition of $T$ at least once and no transition outside
$T$. Finally, we assume that each TELA $\mA$ contains only states $q$
that are reachable from the initial state $q_I$ (i.e., $q=q_I$ or
there is a path from $q_I$ to $q$) as states that are not reachable
from $q_I$ can be eliminated without any impact on $L(\mA)$. We also
assume that each TELA has at least one SCC as automata without any SCC
trivially describe an empty language.

@@

To analyze the length of the formulas, we start with its subformulas.
One can easily check that
$|\msat_\varphi(\vec{e})|\in\bigO(|\varphi|\cdot|\delta|)$ and
$|\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})|\in\bigO(C\cdot
K\cdot|\delta|)$. Further,
$|\mcycle_1(\vec{e})|,|\mcycle_2(\vec{e})|\in\bigO(S\cdot|\delta|)$, % and
%$|\mcycle_2(\vec{e})|\in\bigO(S\cdot|\delta|+|\delta|)=\bigO(S\cdot|\delta|)$,
where $S$ is the number of SCCs in the automaton.  Next,
$|\mcycle_3(\vec{e})|\in\bigO(|\delta|+|Q|)$, which can be simplified
to $|\mcycle_3(\vec{e})|\in\bigO(|\delta|)$ as $|Q|\le|\delta|$
follows from the assumptions that all states are reachable and each
automaton has at least one SCC. Altogether, we get
$|\Phi_{1,C,K}|,|\Phi_{2,C,K}|\in\bigO(S\cdot|\delta|+|\varphi|\cdot|\delta|+C\cdot
K\cdot|\delta|)=\bigO\big((S+|\varphi|+C\cdot K)\cdot|\delta|\big)$
and
$|\Phi_{3,C,K}|\in\bigO(|\delta|+|\varphi|\cdot|\delta|+C\cdot
K\cdot|\delta|)=\bigO\big((|\varphi|+C\cdot K)\cdot|\delta|\big)$.
Note that the formula $\Phi_{3,C,K}$ is asymptotically shorter than
$\Phi_{1,C,K}$ and $\Phi_{2,C,K}$, but it contains an additional quantifier
alternation.

@@

Finally, we mention several three simple optimizations of the formula
construction, which are always applied in the rest of the paper.

@@

The third optimization adds to each formula $\Phi_{j,C,K}$ a conjunct
\[
   \bigwedge_{c\in\{1,2,\ldots,C\}}\bigwedge_{k\in N_K} \neg
   i_{c,k}~\vee~\neg f_{c,k}
\]
saying that any cube does not contain both $\Inf k$ and $\Fin k$ for any
$k$ as such a cube cannot be satisfied by any run.

@@

% \todo{Zminit, ze cela formule
%   ma alepson 1 $\Inf$ nebo $\Fin$? Je to tak, Terko?}

@@

If some call of the function
$\mathit{satisfiable}(\Phi_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In the case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance marks considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
did not decide within the time limit. If $X$ is \texttt{T}, a longer
timeout may lead to further reductions. If the multi-level reduction
is used, the automaton name contains the information from all
levels. For example, the name `\texttt{L1\_5\_U L2\_3\_U L3\_1\_T}'
means that level 1 reduced the number of marks to 6 (reduction to 5 is
impossible on this level), level 2 reduced it to 4, and level 3 to 2
as the QBF solver did not finish in the time limit when trying to
reduce the number of marks to
1.%when trying the automaton may be potentially reduced further.

@@

%\todo{Upravit experimenty: pridat dalsi 2 sady automatu.}
We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. We used the tool
\texttt{ltlfilt} of the Spot library to simplify the formulas and
remove duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinzer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four considered
tools failed on some formulas: it usually reached a timeout of 60
seconds or it produced an automaton that cannot be parsed by the Spot
library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance marks as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators. For each translator, it also provides the numbers of
fails, produced automata with less than two marks, and automata with
at least two marks that are actually used for the experimental
evaluation of our reduction technique.

\begin{table}[t]
  \caption{LTL to automata translators used for the experimental
    evaluation and the numbers of fails and successfully constructed
    automata with less than two marks and with at least two marks.}
\label{tab:ltltranslators}
\centering
% \begin{tabular}{lr@{~~~~~~}lr}
% \toprule
% tool        & version&homepage & ~~~automata \\
% \midrule
% \texttt{ltl2dela}~~~ & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 129\\
% \texttt{ltl2dgra} & 21.0 & & 234\\
% \texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 70\\
% \texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 91\\
% \bottomrule
% \end{tabular}
%
\begin{tabular}{lrrrr}
  \toprule
  &&& \multicolumn{2}{c}{constructed automata with} \\ \cmidrule(lr){4-5}
  tool &\!\!\!version/homepage & fails & 0 or 1 mark & \textbf{at least 2 marks}\\
  \midrule
  \texttt{ltl2dela} & 21.0\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
       & ? & ? & \textbf{129} \\
  \texttt{ltl2dgra} & 21.0\footnotemark[1]
       & ? & ? & \textbf{234}\\
  \texttt{ltl2tgba} & 2.10.4\tablefootnote{\url{https://spot.lrde.epita.fr}}
       & ? & ? & \textbf{70}\\
  \texttt{ltl3tela} & 2.2.0\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
       & ? & ? & \textbf{91}\\
  \bottomrule
\end{tabular}
\end{table}
